Faster generic IND-CCA2 secure KEM using ``encrypt-then-MAC''

### 2024-10-08 etm-transform/manuscript/main.tex
angle more differently

\section{Introduction}\label{sec:introduction}
A key encapsulation mechanism (KEM) is a cryptographic primitive that allows two parties to establish a shared secret over an insecure channel. The combination of KEM for key transport and data encapsulation mechanism (DEM), usually instantiated with authenticated encryption with associated data schemes, form the foundation of today's most widely adopted secure communication protocols such as Transpot Layer Security (TLS) and Secure Shell (SSH).

The accepted security standard for a KEM is \emph{indistinguishability under adaptive chosen-ciphertext attack (IND-CCA2)}. IND-CCA2 security requires that no efficient adversary, with access to a decapsulation oracle throughout the attack, can distinguish a pseudorandom shared secret from truly random noise. However, building an provably IND-CCA2 secure KEM from scratch is immensely difficult. Instead, the most viable approach is to start with a public-key encryption (PKE) scheme with weaker security properties (e.g. OW-CPA or IND-CPA), then put on additional checks for ensuring ciphertext integrity.

One such generic transformation was proposed by Abdalla, Rogaway, and Bellare in 2001 \cite{DBLP:conf/ctrsa/AbdallaBR01}. In its original form, often referred to as ``Hashed ElGamal'', Abdalla's proposal is a hybrid public-key encryption (HPKE) scheme whose chosen-ciphertext security reduces to the strong Diffie-Hellman assumption (that no efficient adversary can violate the computational Diffie-Hellman assumption even with access to a decisional Diffie-Hellman oracle) under the random oracle model.

\subsection{Our contribution}

In this paper, we adapt the ``Hashed ElGamal'' construction to a generic KEM transformation built on top of a PKE, and reduces the IND-CCA2 security of the KEM to the OW-PCA security of the input PKE in the random oracle model. We called our construction the ``encrypt-then-MAC'' KEM transformation due to the conceptual similarity to the namesake symmetric encryption technique for achieving authenticated encryption. A summary of the data flow in the ``encrypt-then-MAC'' KEM can be found in figure \ref{fig:etm-diagram}.

@@

\subsubsection{Security reduction}
From a high level, the ``encrypt-then-MAC'' KEM encapsulates a shared secret by encrypting a randomly sampled PKE plaintext. To ensure the integrity of the ciphertext, a message authenticator is computed on the PKE ciphertext, where the symmetric key is derived by hashing the random PKE plaintext. The PKE ciphertext and message authenticator combine into the KEM ciphertext. For an adversary to produce a valid KEM ciphertext, it must produce a valid tag. If the message authenticator is existentially unforgeable, then the adversary cannot produce a valid tag without knowing the symmetric key, and under the random oracle model, this implies that the adversary must also know the corresonding PKE plaintext, thus rendering the decapsulation oracle's output redundant.

We also extended the discussion of security requirement for the MAC: because each call to the encapsulation routine will use a freshly sampled PKE plaintext, each PKE ciphertext will be signed with a distinct MAC key. In other words, the MAC only needs to be one-time existentially unforgeable, which opens up possibilities of using more efficient constructions than HMAC and CBC-MAC as mentioned in the original paper.

In section \ref{sec:main-results}, we formally state the transformation and prove that the ``encrypt-then-MAC'' KEM is IND-CCA2 secure if the input PKE is OW-PCA secure and the input MAC is one-time existentially unforgeable.

\subsubsection{Performance improvements}
A major advantage of the ``encrypt-then-MAC'' construction is the low performance overhead added to the underlying PKE scheme. We instantiated the ``encrypt-then-MAC'' KEM with the underlying PKE routines of ML-KEM and compared its performance to ML-KEM, which uses the Fujisaki-Okamoto transformation. The Fujisaki-Okamoto transformation uses \emph{de-randomization} and \emph{re-encryption} to achieve chosen ciphertext security, which result in substantial performance penalty. When combined with the Poly1305 message authenticator, our construction $\text{ML-KEM}^+$ achieves on average 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles for encapsulation when compared to ML-KEM.

>>>>

\section{Introduction}\label{sec:introduction}
A key encapsulation mechanism (KEM) \cite{DBLP:journals/iacr/Shoup01} is a cryptographic primitive that allows two parties to establish a shared secret over an insecure channel. The desired security standard for a KEM is called indistinguishability under adaptive chosen-ciphertext attack (IND-CCA2). Intuitively speaking, IND-CCA2 security requires that no efficient adversary can distinguish a pseudorandom shared secret from truly random noise, even with access to a decapsulation oracle throughout the attack. However, building a provably IND-CCA2 KEM is tremendously difficult. Early attempts without formal proof, such as RSA encryption defined in PKCS\#1 v1.5 \cite{DBLP:journals/rfc/rfc2313}, were later shown to be vulnerable to practical chosen-ciphertext attack \cite{DBLP:conf/crypto/Bleichenbacher98}. In recent decades, the most viable approach has been to start with cryptographic primitives possessing weaker security properties, such as a public-key encryption (PKE) scheme with one-way security under chosen-plaintext attack (OW-CPA), then add steps to ensure ciphertext non-malleability \cite{bellare1998relations}. Some of the earliest proposals for generic IND-CCA2 secure constructions include OAEP \cite{bellare1995optimal}, Fujisaki-Okamoto transformation \cite{DBLP:conf/crypto/FujisakiO99}\cite{DBLP:journals/joc/FujisakiO13}, REACT \cite{DBLP:conf/ctrsa/OkamotoP01}, and GEM \cite{DBLP:conf/ctrsa/CoronHJPPT02}.

On the other hand, chosen-ciphertext security is a solve problem in symmetric cryptography. It is well understood that, by combining an IND-CPA secure symmetric encryption scheme with an existentially unforgeable message authentication code (MAC) in a pattern called ``encrypt-then-MAC'' \cite{DBLP:conf/crypto/Krawczyk01}, one can build an authenticated encryption scheme \cite{DBLP:conf/asiacrypt/BellareN00} that achieves IND-CCA2 security. While this technique cannot be directly applied in the context of public-key cryptography due to the lack of a shared symmetric key between the two communicating parties, the concept of authenticating ciphertext using a MAC still has strong merits. Abdalla, Rogaway, and Bellare proposed DHIES (also known as ``Hashed ElGamal'')\cite{DBLP:journals/iacr/AbdallaBR99}\cite{DBLP:conf/ctrsa/AbdallaBR01}, a hybrid public-key encryption (HPKE) scheme whose IND-CCA2 security reduces to the Gap Diffie-Hellman assumption \cite{DBLP:conf/pkc/OkamotoP01} under the random oracle model. The technique behind DHIES is to derive both the shared secret and a symmetric MAC key by hashing a random PKE plaintext, encrypt the PKE plaintext, then authenticate the PKE ciphertext using the previously derived MAC key. Where the Gap Diffie-Hellman assumption holds and the MAC is existentially unforgeable, no efficient adversary can recover the decryption of an unknown ciphertext even with access to a decryption oracle because it cannot produce a valid tag for such unknown ciphertext.

\subsection{Our contribution}
In this paper, we propose a generic IND-CCA2 secure KEM using techniques adapted from DHIES. We call our proposed scheme the ``encrypt-then-MAC'' transformation due to the conceptual similarity to the namesake symmetric encryption technique for achieving authenticated encryption. A summary of the data flow in the ``encrypt-then-MAC'' transformation can be found in figure \ref{fig:etm-diagram}.

@@

Our scheme mainly differs from DHIES in its versatility. Whereas the IND-CCA2 security of DHIES reduces specifically to the Gap Diffie-Hellman assumption, the chosen-ciphertext security of the ``encrypt-then-MAC'' KEM reduces more generally to the OW-PCA security \cite{DBLP:conf/ctrsa/OkamotoP01} of the input scheme (we will show in section \ref{sec:main-results} that the Gap Diffie-Hellman assumption implies a special case of OW-PCA security). In addition, we propose that because each call to encapsulation samples a fresh PKE plaintext, the ``encrypt-then-MAC'' KEM can be instantiated with one-time secure MAC such as Poly1305, which greatly improves the computational efficiency of the scheme.

\subsubsection{Performance improvements}
In section \ref{sec:implementation}, we compare the computational performance in a variety of scenarios between ``encrypt-then-MAC'' KEM to the modular Fujisaki-Okamoto transformation \cite{DBLP:conf/tcc/HofheinzHK17}. A major advantage of the ``encrypt-then-MAC'' construction is the low performance overhead added to the underlying PKE scheme. We instantiated the ``encrypt-then-MAC'' KEM with the underlying PKE routines of ML-KEM and compared its performance to ML-KEM, which uses the Fujisaki-Okamoto transformation. The Fujisaki-Okamoto transformation uses \emph{de-randomization} and \emph{re-encryption} to achieve chosen ciphertext security, which result in substantial performance penalty. When combined with the Poly1305 message authenticator, our construction $\text{ML-KEM}^+$ achieves on average 72\%-80\% reduction of CPU cycles needed for decapsulation while only incurring 2\%-7\% increase of CPU cycles for encapsulation when compared to ML-KEM.
